### 2019-03-25 sections/properties.tex
push fixmes in margin for last proofread
remove last note by Alejandro
rename state s to sigma everywhere (to be compatible with next paper)

Additionally, we show that the normalisation semantics is a big-step semantics. \fixme{Alejandro: ?}
While at first sight, this might seem obvious, the fact that we are dealing with
state complicates matters.

@@

\begin{theorem}[Type preservation under evaluation]
  % For all well typed expressions $e$ and states $s$,
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\tau$ and $\Gamma,\Sigma \infers s$,
  if $e,s \evaluate e',s'$,
  then $\Gamma,\Sigma \infers e':\tau$ and $\Sigma \infers s'$.
  \label{thm:pres-eval}
\end{theorem}

Where $\Sigma \infers s$ means that for all $l\in s$, it holds that
$\Sigma(l)=\Gamma,\Sigma\infers s(l)$.

@@

\begin{theorem}[Type preservation under normalisation]
  % For all well typed expressions $e$ and states $s$,
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $e,s \normalise e',s'$,
  then $\Gamma,\Sigma \infers e':\Task\tau$ and $\Sigma \infers s'$.
  \label{thm:pres-norm}
\end{theorem}

@@

\begin{lemma}[Task value preserves types]
  % For all well typed expressions $e$ and states $s$,
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $\Value{(e,s)}=v$,
  then $v:\tau$.
  \label{lem:presvalue}
\end{lemma}

\begin{lemma}[Striding preserves types]
  For all expressions $e$ and states $s$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $e,s \stride e',s'$,
  then $\Gamma,\Sigma \infers e':\Task\tau$ and $\Sigma \infers s'$.
  \label{thm:pres-stride}
\end{lemma}

@@

\begin{theorem}[Type preservation under handling]
  % For all well typed expressions $e$, states $s$, and inputs $i$,
  For all expressions $e$, states $s$ and inputs $i$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $ e,s \handle{i} e',s'$,
  then $\Gamma,\Sigma\infers e':\Task\tau$ and $\Sigma\infers s'$.
  \label{thm:pres-handle}
\end{theorem}

@@

We define what we mean with progress in \cref{thm:prog-norm}.
\begin{theorem}[Progress under handling]
  For all well typed expressions $e$ and states $s$,
  % For all expressions $e$ and states $s$
  % such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  if $e,s \normalise e',s'$,
  then either $\Failing(e', s')$
  or there exist $e''$, $s''$, and $i$ such that $e',s'\handle{i} e'',s''$.
  \label{thm:prog-norm}
\end{theorem}

Where a well typed expression $e$ means that $\Gamma,\Sigma\infers e:\tau$ for
some type $\tau$, and a well typed state means that $\Sigma\infers s$.

If an expression $e$ and state $s$ are well-typed, then after normalisation, the pair $e',s'$
either fails, or there exists some input $i$ that can be handled by it under the handling semantics.
In order to prove this Theorem, we require an additional theorem.

@@

\begin{theorem}[Failing means no interaction possible]
  For all well typed expressions $e$ and states $s$,
  % For all expressions $e$ and states $s$
  % such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers s$,
  and $e,s \normalise e',s'$,
  we have that $\Failing(e',s') = \True$,
  if and only if there is no input $i$
  such that $e',s'\handle{i} e'',s''$ for some $e''$ and $s''$.
  \label{thm:failing}
\end{theorem}

The Theorem above states that an expression $e$ and state $s$ are failing, if,
after normalisation, there exists no input that can be handled by it.
We prove the theorem to be true by induction on $e'$.

@@

\begin{proof}
  Given $\Gamma,\Sigma\infers e:\Task\tau$ and $\Sigma\infers s$ and after
  normalisation $e,s \normalise e',s'$, we find ourselves in either one of the
  following situations:

  There exists an $i$ such that $e',s'\handle{i} e'',s''$.

  There does not exist an $i$ such that $e',s'\handle{i} e'',s''$. In this case, we
  know that $\Failing(e',s')$ must be true, by \cref{thm:failing}.
\end{proof}

@@

\begin{theorem}[Inputs function is sound and complete]
  For all well typed expressions $e$, states $s$, and inputs $i$,
  % For all expressions $e$, states $s$, and inputs $i$
  % such that $\Gamma,\Sigma \infers e:\tau$ and $\Sigma \infers s$,
  we have that $i \in \Inputs{(e,s)}$ if and only if $e,s \handle{i} e',s'$.
  \label{thm:safety-i}
\end{theorem}

>>>>

Additionally, we show that the normalisation semantics is a big-step semantics.
%XXX: Alejandro: ?
% Niks meer mee doen, was voor PLDI reviewers duidelijk genoeg. --TS
While at first sight, this might seem obvious, the fact that we are dealing with
state complicates matters.

@@

\begin{theorem}[Type preservation under evaluation]
  % For all well typed expressions $e$ and states $\sigma$,
  For all expressions $e$ and states $\sigma$
  such that $\Gamma,\Sigma \infers e:\tau$ and $\Gamma,\Sigma \infers \sigma$,
  if $e,\sigma \evaluate e',\sigma'$,
  then $\Gamma,\Sigma \infers e':\tau$ and $\Sigma \infers \sigma'$.
  \label{thm:pres-eval}
\end{theorem}

Where $\Sigma \infers \sigma$ means that for all $l\in \sigma$, it holds that
$\Sigma(l)=\Gamma,\Sigma\infers \sigma(l)$.

@@

\begin{theorem}[Type preservation under normalisation]
  % For all well typed expressions $e$ and states $\sigma$,
  For all expressions $e$ and states $\sigma$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  if $e,\sigma \normalise e',\sigma'$,
  then $\Gamma,\Sigma \infers e':\Task\tau$ and $\Sigma \infers \sigma'$.
  \label{thm:pres-norm}
\end{theorem}

@@

\begin{lemma}[Task value preserves types]
  % For all well typed expressions $e$ and states $\sigma$,
  For all expressions $e$ and states $\sigma$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  if $\Value{(e,\sigma)}=v$,
  then $v:\tau$.
  \label{lem:presvalue}
\end{lemma}

\begin{lemma}[Striding preserves types]
  For all expressions $e$ and states $\sigma$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  if $e,\sigma \stride e',\sigma'$,
  then $\Gamma,\Sigma \infers e':\Task\tau$ and $\Sigma \infers \sigma'$.
  \label{thm:pres-stride}
\end{lemma}

@@

\begin{theorem}[Type preservation under handling]
  % For all well typed expressions $e$, states $\sigma$, and inputs $i$,
  For all expressions $e$, states $\sigma$ and inputs $i$
  such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  if $ e,\sigma \handle{i} e',\sigma'$,
  then $\Gamma,\Sigma\infers e':\Task\tau$ and $\Sigma\infers \sigma'$.
  \label{thm:pres-handle}
\end{theorem}

@@

We define what we mean with progress in \cref{thm:prog-norm}.
\begin{theorem}[Progress under handling]
  For all well typed expressions $e$ and states $\sigma$,
  % For all expressions $e$ and states $\sigma$
  % such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  if $e,\sigma \normalise e',\sigma'$,
  then either $\Failing(e', \sigma')$
  or there exist $e''$, $\sigma''$, and $i$ such that $e',\sigma'\handle{i} e'',\sigma''$.
  \label{thm:prog-norm}
\end{theorem}

Where a well typed expression $e$ means that $\Gamma,\Sigma\infers e:\tau$ for
some type $\tau$, and a well typed state means that $\Sigma\infers \sigma$.

If an expression $e$ and state $\sigma$ are well-typed, then after normalisation, the pair $e',\sigma'$
either fails, or there exists some input $i$ that can be handled by it under the handling semantics.
In order to prove this Theorem, we require an additional theorem.

@@

\begin{theorem}[Failing means no interaction possible]
  For all well typed expressions $e$ and states $\sigma$,
  % For all expressions $e$ and states $\sigma$
  % such that $\Gamma,\Sigma \infers e:\Task\tau$ and $\Sigma \infers \sigma$,
  and $e,\sigma \normalise e',\sigma'$,
  we have that $\Failing(e',\sigma') = \True$,
  if and only if there is no input $i$
  such that $e',\sigma'\handle{i} e'',\sigma''$ for some $e''$ and $\sigma''$.
  \label{thm:failing}
\end{theorem}

The Theorem above states that an expression $e$ and state $\sigma$ are failing, if,
after normalisation, there exists no input that can be handled by it.
We prove the theorem to be true by induction on $e'$.

@@

\begin{proof}
  Given $\Gamma,\Sigma\infers e:\Task\tau$ and $\Sigma\infers \sigma$ and after
  normalisation $e,\sigma \normalise e',\sigma'$, we find ourselves in either one of the
  following situations:

  There exists an $i$ such that $e',\sigma'\handle{i} e'',\sigma''$.

  There does not exist an $i$ such that $e',\sigma'\handle{i} e'',\sigma''$. In this case, we
  know that $\Failing(e',\sigma')$ must be true, by \cref{thm:failing}.
\end{proof}

@@

\begin{theorem}[Inputs function is sound and complete]
  For all well typed expressions $e$, states $\sigma$, and inputs $i$,
  % For all expressions $e$, states $\sigma$, and inputs $i$
  % such that $\Gamma,\Sigma \infers e:\tau$ and $\Sigma \infers \sigma$,
  we have that $i \in \Inputs{(e,\sigma)}$ if and only if $e,\sigma \handle{i} e',\sigma'$.
  \label{thm:safety-i}
\end{theorem}

### 2019-03-25 sections/semantics.tex
rename state s to sigma everywhere (to be compatible with next paper)

The host language evaluates expressions using a big-step semantics.
Evaluating an expression $e$ in state $s$ into a value $v$ in state $s'$ is denoted by $\RelationE$.
To ease reasoning about references, we choose a call-by-value evaluation strategy.

@@

The normalisation semantics is responsible for reducing expressions of type $\Task$ until they are ready to handle input.
It is a big-step semantics, and makes use of evaluation of the host language.
We write $\RelationN$ to describe that
an expression $e$ in state $s$ normalises to task $t$ in state $s'$.

@@

The split between striding and normalisation is due to mutable references.
Consider the following example, where $s=\set{l\mapsto\False}$.
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (l := True; edit <<>>)
\end{TASK}
\pagebreak[2]
\refrule{S-And} reduces this expression in one step to
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (edit <<>>)
\end{TASK}
with $s'=\set{l\mapsto\True}$.
This expression is not normalised, because the left task can take a step.
The issue here lies in the fact that the right task updates $l$.
To overcome this problem, the \refrule{N-Done} and \refrule{N-Repeat} rules ensure that striding is applied until the state $s$ becomes stable and no further normalisation can take place.
%XXX: Alejandro: The example is not 100\% clear to me

The striding semantics is responsible for reducing internal steps and internal choices.
A stride from task $t$ in state $s$ to $t'$ in state $s'$ is denoted by $\RelationS$.
The rules for striding are given in \cref{fig:striding-semantics}.
Tasks like editors, fail and external choice are not further reduced.
For external choice and parallel there are congruence rules.

@@

Handling input is done by the \emph{handling} semantics shown in \cref{fig:handling-semantics}.
It is a small step semantics with labelled transitions.
It takes a task $t$ in a state $s$ and an input $i$, and yields a new task $t'$ in a new state $s'$.

>>>>

The host language evaluates expressions using a big-step semantics.
Evaluating an expression $e$ in state $\sigma$ into a value $v$ in state $\sigma'$ is denoted by $\RelationE$.
To ease reasoning about references, we choose a call-by-value evaluation strategy.

@@

The normalisation semantics is responsible for reducing expressions of type $\Task$ until they are ready to handle input.
It is a big-step semantics, and makes use of evaluation of the host language.
We write $\RelationN$ to describe that
an expression $e$ in state $\sigma$ normalises to task $t$ in state $\sigma'$.

@@

The split between striding and normalisation is due to mutable references.
Consider the following example, where $\sigma=\set{l\mapsto\False}$.
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (l := True; edit <<>>)
\end{TASK}
\pagebreak[2]
\refrule{S-And} reduces this expression in one step to
\begin{TASK}
  (update l >>= \x:Bool. if x then e else fail) <&> (edit <<>>)
\end{TASK}
with $\sigma'=\set{l\mapsto\True}$.
This expression is not normalised, because the left task can take a step.
The issue here lies in the fact that the right task updates $l$.
To overcome this problem, the \refrule{N-Done} and \refrule{N-Repeat} rules ensure that striding is applied until the state $\sigma$ becomes stable and no further normalisation can take place.
%XXX: Alejandro: The example is not 100\% clear to me

The striding semantics is responsible for reducing internal steps and internal choices.
A stride from task $t$ in state $\sigma$ to $t'$ in state $\sigma'$ is denoted by $\RelationS$.
The rules for striding are given in \cref{fig:striding-semantics}.
Tasks like editors, fail and external choice are not further reduced.
For external choice and parallel there are congruence rules.

@@

Handling input is done by the \emph{handling} semantics shown in \cref{fig:handling-semantics}.
It is a small step semantics with labelled transitions.
It takes a task $t$ in a state $\sigma$ and an input $i$, and yields a new task $t'$ in a new state $\sigma'$.
